---
title: "Multilevel Modelling & MAIHDA"
subtitle: "From Educational Segregation to Intersectional Analysis"
author: 
  - name: "Dr Yiyang Gao"
    affiliations:
      - "PhD in Advanced Quantitative Methods, Q-Step Centre, Exeter"
      - "Current Postdoc, Evidence Centre for Education, Durham University"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    show-slide-number: all
    transition: slide
    background-transition: fade
    width: 1024
    height: 768
    margin: 0.1
    center: true
    preview-links: auto
    logo: ""
    footer: "MAIHDA: Three Research Applications"
    css: custom.css
---

```{r setup, include=FALSE}
# Load required packages
library(ggplot2)
library(dplyr)
library(tidyr)
# Note: ggforce is only needed if you want to use geom_circle
# If not installed, you can comment out the spatial visualization plot
```

## Research Profile {background-color="#7B3F99"}

:::: {.columns}

::: {.column width="33%"}
### 🏫 Ethnicity & Segregation

**Evolving Patterns of Ethnic Segregation in English Schools**  
*Population, Space and Place (R&R)*

**⭐ Multilevel Analysis of Ethnic Clustering Across Local Schools**  
*British Education Research Journal (R&R)*  
*★ Featured in this presentation*
:::

::: {.column width="33%"}
### 💼 Workforce Inequality

**⭐ Understanding Teacher Retention in England: A Multi-Cohort Survival Analysis**  
*Research Papers in Education (Under Review)*  
*★ Featured in this presentation*

**Spatial-Temporal Patterns of Teacher Career Progression**  
*Presenting at: ADR UK & RGS-IBG*
:::

::: {.column width="34%"}
### 🗺️ Accessibility, Simulation, Health

**Convenor: "Systems and Simulation Approaches in Population Research"**  
*BSPS | Collaboration with Glasgow*

**⭐ Grant: Educational Equity - Transport & School Segregation**  
*With Department of Sociology, Durham*  
*★ Featured in this presentation*
:::

::::

::: {.notes}
Research positioning at the intersection of Quantitative Geography (Space), Demography/Sociology (Population), Education, and Health
:::

---

## What is MAIHDA?

### Multilevel Analysis of Individual Heterogeneity and Discriminatory Accuracy

::: {.callout-important}
## The "Gold Standard" for Quantitative Intersectional Analysis
Treats intersectional identities as random effects in multilevel models, revealing how multiple characteristics combine to create unique experiences beyond simple additive effects
:::

:::: {.columns}

::: {.column width="50%"}
### Conventional Approach:
- Examines factors separately
- Assumes additive effects
- Misses critical intersections
- Generic policy recommendations
:::

::: {.column width="50%"}
### MAIHDA Approach:
- Creates intersectional strata
- Reveals multiplicative effects
- Handles small cell sizes via shrinkage
- Enables targeted interventions
:::

::::

::: {.callout-warning}
## Three Extended MAIHDA Applications:
- **Spatial MAIHDA** → School Segregation Study
- **Longitudinal MAIHDA** → Teacher Retention Study
- **Policy Evaluation MAIHDA** → Transport & Accessibility Study
:::

---

## Study 1: Ethnic School Segregation - Current Approach

:::: {.columns}

::: {.column width="50%"}
### Research Question
How do ethnic clustering patterns vary across schools, local authorities, and over time in England?

### Current Method
- Three-level multilevel model
- Years → Local Authorities → Regions
- Observation unit: Segregation scores (LA-year level)
- Entropy-based segregation index
- Examines factors separately

```
Y_tij = γ₀₀₀ + β₁(Selective)_ij + β₂(Academies)_ij 
        + β₃(Faith)_ij + β₄(% ethnic)_ij 
        + u₀₀j + r₀ij + e_tij
```
:::

::: {.column width="50%"}
### Key Finding
**Counterintuitive: Higher ethnic % = Lower concentration**

```{r}
#| echo: false
#| fig-height: 4
library(ggplot2)
library(dplyr)

variance_data <- data.frame(
  Group = rep(c("Bangladeshi", "Indian", "White British"), each = 4),
  Component = rep(c("Year", "Local Authority", "Region", "Residual"), 3),
  Percentage = c(30, 41, 26, 3,  # Bangladeshi
                 44, 21, 26, 9,  # Indian
                 61, 31, 5, 3)    # White British
)

variance_data$Component <- factor(variance_data$Component, 
                                  levels = c("Residual", "Region", "Local Authority", "Year"))

ggplot(variance_data, aes(x = Group, y = Percentage, fill = Component)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("Residual" = "#D3D3D3", 
                               "Region" = "#98FB98", 
                               "Local Authority" = "#87CEEB", 
                               "Year" = "#FFB6C1")) +
  labs(y = "Variance Explained (%)", x = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("Variance Components by Ethnic Group")
```
:::

::::

::: {.callout-note}
## Limitation
Cannot explain WHY this happens differently for different groups or WHERE spatial clusters of disadvantage exist
:::

---

## Study 1: Spatial MAIHDA Enhancement

:::: {.columns}

::: {.column width="50%"}
### Method Enhancement:
- 48 intersectional strata: Ethnicity × SES × School Type
- Cross-classified with spatial units (LSOAs)
- Spatial weights matrix for autocorrelation
- Monte Carlo uncertainty propagation (10,000 simulations)

```
Y_ijk = γ₀ + (1|strata_j) + (1|spatial_k) 
        + W*spatial_effects + ε_ijk
```

### Simulated Results:

| Intersectional Group | Segregation Index | Spatial Clustering |
|---------------------|-------------------|-------------------|
| Pakistani × Low SES × Faith School | **142.5** | Moran's I = 0.82 |
| White British × High SES × Academy | 45.2 | Moran's I = 0.21 |
| Black × Low SES × Community | 118.3 | Moran's I = 0.67 |
:::

::: {.column width="50%"}
### Monte Carlo Spatial Simulation

```{r}
#| echo: false
#| fig-height: 5
library(ggplot2)

# Create spatial cluster visualization
set.seed(123)

# If ggforce is available, use geom_circle; otherwise use geom_point
if (requireNamespace("ggforce", quietly = TRUE)) {
  library(ggforce)
  clusters <- data.frame(
    x = c(2, 2.5, 2.3, 6, 6.5, 9, 9.3),
    y = c(3, 3.3, 2.5, 4, 3.8, 3.5, 2.8),
    size = c(60, 45, 35, 30, 25, 50, 40),
    type = c("High", "High", "High", "Low", "Low", "Moderate", "Moderate"),
    label = c("High Segregation\nPakistani×LowSES", "", "", 
              "Low Segregation", "", "Moderate", "")
  )
  
  ggplot(clusters) +
    geom_circle(aes(x0 = x, y0 = y, r = size/20, fill = type), alpha = 0.6) +
    geom_text(data = clusters[clusters$label != "",], 
              aes(x = x, y = y, label = label), size = 3) +
    geom_curve(aes(x = 3.5, y = 3.2, xend = 5.5, yend = 3.8),
               arrow = arrow(length = unit(0.3, "cm")),
               curvature = -0.3) +
    annotate("text", x = 4.5, y = 4.2, label = "Spatial Spillover", size = 3) +
    scale_fill_manual(values = c("High" = "#FF6B6B", 
                                 "Low" = "#4ECDC4", 
                                 "Moderate" = "#FFE66D")) +
    theme_minimal() +
    theme(legend.position = "none") +
    coord_fixed() +
    xlim(0, 11) + ylim(1, 5) +
    labs(title = "Spatial Clustering of School Segregation")
} else {
  # Alternative visualization without ggforce
  clusters <- data.frame(
    x = c(2, 2.5, 2.3, 6, 6.5, 9, 9.3),
    y = c(3, 3.3, 2.5, 4, 3.8, 3.5, 2.8),
    size = c(60, 45, 35, 30, 25, 50, 40),
    type = c("High", "High", "High", "Low", "Low", "Moderate", "Moderate"),
    label = c("High Segregation\nPakistani×LowSES", "", "", 
              "Low Segregation", "", "Moderate", "")
  )
  
  ggplot(clusters) +
    geom_point(aes(x = x, y = y, size = size, color = type), alpha = 0.6) +
    geom_text(data = clusters[clusters$label != "",], 
              aes(x = x, y = y, label = label), size = 3) +
    geom_curve(aes(x = 3.5, y = 3.2, xend = 5.5, yend = 3.8),
               arrow = arrow(length = unit(0.3, "cm")),
               curvature = -0.3) +
    annotate("text", x = 4.5, y = 4.2, label = "Spatial Spillover", size = 3) +
    scale_color_manual(values = c("High" = "#FF6B6B", 
                                  "Low" = "#4ECDC4", 
                                  "Moderate" = "#FFE66D")) +
    scale_size_continuous(range = c(5, 15), guide = "none") +
    theme_minimal() +
    theme(legend.position = "none") +
    xlim(0, 11) + ylim(1, 5) +
    labs(title = "Spatial Clustering of School Segregation")
}
```
:::

::::

::: {.callout-important}
## Key Finding
Monte Carlo reveals 95% CI [135.2, 149.8] for Pakistani×LowSES×Faith clusters. Spatial autocorrelation shows segregation "hot spots" requiring targeted intervention in specific neighborhoods, not just schools.
:::

---

## Study 2: Teacher Retention - Current Approach

:::: {.columns}

::: {.column width="50%"}
### Current Method:
- Cox proportional hazards model
- 650,000 teachers tracked
- Shows WHO leaves and WHEN
- Cannot explain intersectional vulnerabilities

### Key Statistics:

| Group | 11-Year Retention |
|-------|-------------------|
| Female | 16.54% |
| Male | 12.96% |
| With ITT | 16.85% |
| No ITT | 9.75% |
| White British | 15.88% |
| Ethnic Minority | 13.50% |
:::

::: {.column width="50%"}
### Kaplan-Meier Survival Curves:

```{r}
#| echo: false
#| fig-height: 5
library(ggplot2)
library(survival)

# Simulate survival data
time <- seq(0, 11, 0.1)
itt_survival <- exp(-0.15 * time)
no_itt_survival <- exp(-0.25 * time)

survival_data <- data.frame(
  time = rep(time, 2),
  survival = c(itt_survival, no_itt_survival),
  group = rep(c("With ITT (16.85%)", "No ITT (9.75%)"), each = length(time))
)

ggplot(survival_data, aes(x = time, y = survival, color = group)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("With ITT (16.85%)" = "#FF6B6B", 
                                "No ITT (9.75%)" = "#4ECDC4")) +
  labs(x = "Years in Teaching", 
       y = "Retention Rate",
       title = "Teacher Retention by Initial Teacher Training Status") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  scale_y_continuous(labels = scales::percent)
```
:::

::::

::: {.callout-note}
## Limitation
Treats characteristics separately - misses critical intersectional vulnerabilities like "Black male teachers without ITT in London"
:::

---

## Study 2: Longitudinal MAIHDA Enhancement

:::: {.columns}

::: {.column width="50%"}
### Method Enhancement:
- 200 intersectional strata: Ethnicity × Gender × ITT × Region
- Time-varying hazard functions
- Monte Carlo trajectory simulation (5,000 paths)
- Identifies critical career junctures

### Intersectional Groups (examples):
- White British × Female × ITT × North
- Black × Male × Non-ITT × London
- Pakistani × Female × ITT × Yorkshire
- Chinese × Male × ITT × Southeast
:::

::: {.column width="50%"}
### Monte Carlo Simulated Trajectories:

```{r}
#| echo: false
#| fig-height: 5
library(ggplot2)

# Simulate trajectory data with confidence bands
time <- seq(0, 11, 0.1)
set.seed(456)

# Group 1: White × Female × ITT × North
traj1_mean <- exp(-0.12 * time)
traj1_upper <- exp(-0.10 * time)
traj1_lower <- exp(-0.14 * time)

# Group 2: Black × Male × Non-ITT × London
traj2_mean <- exp(-0.28 * time)
traj2_upper <- exp(-0.25 * time)
traj2_lower <- exp(-0.31 * time)

# Create separate data frames for ribbons
ribbon1 <- data.frame(
  time = time,
  mean = traj1_mean,
  upper = traj1_upper,
  lower = traj1_lower,
  group = "White × Female × ITT × North"
)

ribbon2 <- data.frame(
  time = time,
  mean = traj2_mean,
  upper = traj2_upper,
  lower = traj2_lower,
  group = "Black × Male × Non-ITT × London"
)

# Combine for line plot
line_data <- rbind(
  data.frame(time = time, value = traj1_mean, group = "White × Female × ITT × North"),
  data.frame(time = time, value = traj2_mean, group = "Black × Male × Non-ITT × London")
)

ggplot() +
  geom_ribbon(data = ribbon1, aes(x = time, ymin = lower, ymax = upper), 
              fill = "#7B3F99", alpha = 0.2) +
  geom_ribbon(data = ribbon2, aes(x = time, ymin = lower, ymax = upper), 
              fill = "#FF6B6B", alpha = 0.2) +
  geom_line(data = line_data, aes(x = time, y = value, color = group), 
            linewidth = 1.5) +
  geom_point(x = 2, y = exp(-0.28 * 2), color = "#FFD700", size = 4) +
  annotate("text", x = 2.5, y = exp(-0.28 * 2) + 0.05, 
           label = "Year 2 Crisis", size = 3) +
  scale_color_manual(values = c("White × Female × ITT × North" = "#7B3F99", 
                                "Black × Male × Non-ITT × London" = "#FF6B6B")) +
  labs(x = "Years", y = "Retention Probability",
       title = "Career Trajectories with 95% Confidence Intervals") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```
:::

::::

::: {.callout-important}
## Critical Finding
Monte Carlo reveals Black male non-ITT teachers in London face accelerated attrition at Year 2 (HR jumps to 2.8, 95% CrI: [2.1, 3.7]). This specific vulnerability was invisible in conventional analysis.
:::

---

## Study 3: Transport & Educational Access - Current Approach

:::: {.columns}

::: {.column width="50%"}
### Research Question:
How does public transport infrastructure affect school segregation and educational equity?

### Current Planned Method:
- Agent-based modeling (ABM)
- MATSim transport simulation
- Quality-Accessibility Index
- Compare London (free transport) vs Leeds

### Focus:
Simulate household school choices under different transport scenarios
:::

::: {.column width="50%"}
### Preliminary Transport Barriers:

::: {.callout-note}
## Average Journey Times to Quality Schools:
- Low Income Households: **47 min**
- High Income Households: **18 min**
- No Car Access: **52 min**
- Car Access: **15 min**
:::

::: {.callout-warning}
## Key Policy Question:
Would free bus passes (like London) reduce educational inequality?
:::
:::

::::

::: {.callout-note}
## Limitation
Current approach treats demographic groups separately - misses how transport barriers compound for specific intersectional groups
:::

---

## Study 3: Policy Evaluation MAIHDA Enhancement

:::: {.columns}

::: {.column width="50%"}
### Method Enhancement:
- 60 intersectional strata: Ethnicity × Income × Car Access × Zone
- Random slopes for policy effects
- Monte Carlo policy simulation (10,000 scenarios)
- Pre/post free transport comparison

```
Y_ij = γ₀ + β₁(Policy) + u₁j(Policy×Strata) 
       + (1|strata_j) + ε_ij
```

### Policy Scenarios:
1. Free bus passes for all students
2. Targeted passes for low-income areas
3. Enhanced bus frequency
4. New routes to quality schools
:::

::: {.column width="50%"}
### Monte Carlo Policy Impact Simulation:

| Intersectional Group | Current Access | Post-Policy | % Change |
|---------------------|----------------|-------------|----------|
| Pakistani × Low Income × No Car | **2.3 schools** | 5.8 schools | +152% |
| White × High Income × Car | 8.7 schools | 9.1 schools | +5% |
| Black × Low Income × No Car | **3.1 schools** | 6.2 schools | +100% |
| Chinese × Mid Income × Car | 6.4 schools | 7.0 schools | +9% |

```{r}
#| echo: false
#| fig-height: 3.5
library(ggplot2)
library(tidyr)

policy_data <- data.frame(
  Group = c("Pak×Low×NoCar", "White×High×Car", "Black×Low×NoCar", "Chi×Mid×Car"),
  Pre_Policy = c(2.3, 8.7, 3.1, 6.4),
  Post_Policy = c(5.8, 9.1, 6.2, 7.0)
)

policy_long <- pivot_longer(policy_data, cols = c(Pre_Policy, Post_Policy), 
                            names_to = "Period", values_to = "Schools")

ggplot(policy_long, aes(x = Group, y = Schools, fill = Period)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Pre_Policy" = "#FF6B6B", "Post_Policy" = "#4ECDC4"),
                    labels = c("Pre-Policy", "Post-Policy")) +
  labs(y = "Number of Accessible Schools", x = "",
       title = "Policy Impact by Intersectional Group") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

::::

::: {.callout-important}
## Policy Finding
Monte Carlo reveals 95% CI for impact: Pakistani low-income no-car families gain [4.8, 6.8] schools. Free transport reduces inequality by 42% [38%, 46%] - but only for specific intersectional groups.
:::

---

## MAIHDA: Transforming Understanding of Educational Inequalities {background-color="#7B3F99"}

### Evolution Across Three Studies

| Study | Current Method | MAIHDA Extension | Key Innovation |
|-------|----------------|------------------|----------------|
| **School Segregation** | Conventional Multilevel | Spatial MAIHDA | Reveals neighborhood clusters of intersectional disadvantage |
| **Teacher Retention** | Survival Analysis | Longitudinal MAIHDA | Identifies career crisis points for specific groups |
| **Transport Access** | Agent-Based Model | Policy Evaluation MAIHDA | Quantifies differential policy impacts |

:::: {.columns}

::: {.column width="50%"}
### Methodological Innovation:
- Extending MAIHDA to spatial, longitudinal, and policy domains
- Monte Carlo uncertainty quantification
- Large dataset expertise (NPD, SWC)
:::

::: {.column width="50%"}
### Why MAIHDA Matters:
- Identifies WHO needs help (intersections)
- Shows WHERE to intervene (spatial)
- Reveals WHEN to act (temporal)
- Quantifies WHAT WORKS (policy)
:::

::::

::: {.callout-important}
## Vision
Advancing intersectional quantitative methods to reveal not just that inequalities exist, but precisely **where**, **for whom**, **when**, and **why** – transforming our ability to design interventions that actually work
:::